Synthesizing Symmetric Lenses

### 2019-01-28 papers/SymmetricLenses/main.tex
Polishing

\noindent
In today's big-data world, similar information is often
stored in different places and in different formats. For instance, electronic
calendars come in iCalendar, vCalendar, and h-event
formats~\cite{calendar-formats}; US taxation data can be transmitted via Tax
XML (used by the US government for e-filing) or TXF (used by
TurboTax)~\cite{tax-formats}; and sewing machines can use embroidery files
in formats with names like PES, DST, and several
others~\cite{embroidery-formats}.
%
\bcp{I'm not satisfied with the embroidery example: It doesn't obviously
  correspond to a bidirectional situation (the sewing machine does not need
  to send back the embroidery file to anybody).  The tax example could be
  criticized on the same grounds.}
%
\bcp{Also, it would be nice to be able to give a quick example of a truly
  symmetric case---do any of these examples work for that?}\afm{h-event and
  vCalendar are truly symmetric. Even the fields related to location alone
  are truly symmetric (though it is subtle to see this).}

One convenient way to maintain consistency between such varied data formats
is to use a \emph{lens}~\cite{Focal2005-long}---a bidirectional program that
can transform updates to data represented in a format $S$ to another format
$T$ and vice versa, providing ``round tripping'' guarantees that help ensure
information is not lost or corrupted as it is transformed back and forth
between different representations.

Domain-specific languages for writing lenses can facilitate building
principled data transformations,
\textit{Synthesizing} lenses 
can make the task even easier.  Indeed, lens languages offer excellent
targets for search-based synthesis, since their specialized, sub-Turing-complete syntax
and expressive type systems drastically
limit the space of possible programs. 
%
Recently, Optician~\cite{optician} has exploited this observation, demonstrating that
it is feasible to synthesize quite complex examples, albeit only for a
restricted class of lenses---so-called bijective lenses. Specifically, given 
source and target formats plus a small number of
examples of corresponding data, the bijective synthesis algorithm will find a
bijective lens matching the examples, if one exists. Often,
it can find an appropriate lens in less than a second with
no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis is so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between related data formats. 
Hence, if the synthesis algorithm can
find any bijection at all, it is likely to be the right one.

However, many real-world data formats share just part of their information
content with other representations.  For instance, two related
database tables might share four out of five columns, but each might have a
different fifth column. 
Or, one ad-hoc system configuration file might include some format-specific meta
data, such as a date or a reference number, whereas the same configuration
file on another operating system does not. 
Indeed among the 48 benchmark problems described in Section \ref{sec:evaluation},
all of the benchmarks taken from Flash Fill~\cite{flashfill} and many of the
benchmarks that synchronize between two ad hoc file formats have this
characteristic, 13 benchmarks in total. Consequently, synthesizing only strict
bijections is often not useful in practice.

Could we extend synthesis tools to more kinds of lenses?  In particular,
what about \emph{symmetric lenses}~\cite{symmetric-lenses}, a class that
includes bijective lenses, ``asymmetric'' lenses where the transformation
from $S$ to $T$ can throw away information that must then be restored when
returning from $T$ to $S$, and even more flexibile transformations that
allow each side to retain information not present in the other?

>>>>

\noindent
In today's data-rich world, similar information is often
stored in different places and in different formats. For instance, electronic
calendars come in iCalendar, vCalendar, and h-event
formats~\cite{calendar-formats}; US taxation data can be transmitted via Tax
XML (used by the US government for e-filing) or TXF (used by
TurboTax)~\cite{tax-formats}; and sewing machines can use embroidery files
in formats with names like PES, DST, and several
others~\cite{embroidery-formats}.
%
\bcp{I'm not satisfied with the embroidery example: It doesn't obviously
  correspond to a bidirectional situation (the sewing machine does not need
  to send back the embroidery file to anybody).  The tax example could be
  criticized on the same grounds.}
%
\bcp{Also, it would be nice to be able to give a quick example of a truly
  symmetric case---do any of these examples work for that?}\afm{h-event and
  vCalendar are truly symmetric. Even the fields related to location alone
  are truly symmetric (though it is subtle to see this).}

One convenient way to maintain consistency between such varied data formats
is to use a \emph{lens}~\cite{Focal2005-long}---a bidirectional program that
can transform updates to data represented in a format $S$ to another format
$T$ and vice versa, providing ``round tripping'' guarantees that help ensure
information is not lost or corrupted as it is transformed back and forth
between different representations.
Domain-specific languages for writing lenses can facilitate building
principled data transformations.

\textit{Synthesizing} lenses 
can make the task even easier.  Indeed, lens languages are excellent
targets for search-based synthesis, since their specialized, sub-Turing-complete syntax
and expressive type systems drastically
limit the space of possible programs. 
%
In particular, the Optician~\cite{optician} system has demonstrated that
it is feasible to synthesize quite complex examples, albeit only for a
restricted class of lenses---so-called bijective lenses. Specifically, given 
source and target formats plus a small number of
examples of corresponding data, Optician's synthesis algorithm is guaranteed
to find a bijective lens matching the examples, if one exists. Indeed, 
it can often find an appropriate lens in less than a second with
no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm can
find any bijection at all, it is fairly likely to be the intended one.

However, the set of real-world use cases for bijective lenses is also
somewhat limited.  A more common situation is where two real-world data
formats share just {\em some} of their information content.  For instance,
two related database tables might share four out of five columns, but each
might have a different fifth column.  Or, one ad-hoc system configuration
file might include some format-specific meta data, such as a date or a
reference number, whereas the same configuration file on another operating
system does not.  Indeed among the 48 benchmark problems described in
Section \ref{sec:evaluation} below, all of the benchmarks taken from Flash
Fill~\cite{flashfill} and many of the benchmarks that synchronize between
two ad hoc file formats have this characteristic, 13 benchmarks in
total.

Can Optician's basic synthesis procedure be extended to a richer class of
lenses? In particular, what about \emph{symmetric
  lenses}~\cite{symmetric-lenses}---a much larger class that includes
bijective lenses, ``asymmetric'' lenses (where the transformation from a
source structure $S$ to a target $T$ can throw away information that must
then be restored when transferring from $T$ to $S$), and even more flexibile
transformations that allow each side to retain information not present in
the other?
